Prefix-free graphs and suffix array construction in sublinear space

### 2023-06-26 manuscript/conclusion.tex
finished

\section{Conclusion}
We introduced the prefix-free graph as a standalone data structure in pangenomics and showed how to use it to build a suffix array of a pangenome in sublinear space.
There are several possible directions for future research on prefix-free graphs.
The first direction may be to answer the question of which additional stringological algorithms and data structures can use prefix-free graphs to improve their applicability to such vast and repetitive datasets as pangenomic data.
Next, the choice of trigger words is, so far, mostly undiscovered territory.
From a computational perspective, the set of trigger words minimizing the size of a resulting prefix-free graph is desirable since it would allow the analysis of larger datasets.
Furthermore, trigger words offer great flexibility in the construction process.
We hypothesize this flexibility could be used for the construction of prefix-free graphs based on biologically significant strings.
In our experiments, we used the stop codon sequences as the trigger words, but other motives, such as different binding sites, recombination hotspots, or repetitive elements, are possible and could better capture some biological phenomena.

>>>>

\section{Conclusion}
We introduced the prefix-free graph as a standalone data structure in pangenomics and showed how to use it to build a suffix array of a pangenome in sublinear space.
There are several possible directions for future research on prefix-free graphs.
The first direction may be to answer the question of which additional stringological algorithms and data structures can use prefix-free graphs to improve their applicability to such vast and repetitive datasets as pangenomic data.
Next, the choice of trigger words is, so far, mostly undiscovered territory.
From a computational perspective, the set of trigger words minimizing the size of a resulting prefix-free graph is desirable since it would allow the analysis of larger datasets.
Furthermore, trigger words offer great flexibility in the construction process.
We hypothesize this flexibility could be used for the construction of prefix-free graphs based on biologically significant strings.
In our experiments, we used the stop codon sequences as the trigger words, but other motives, such as different binding sites, recombination hotspots, or repetitive elements, are possible and could be used to capture particular biological phenomena.

### 2023-06-26 manuscript/main.tex
finished?
finished

\begin{acknowledgments}
    Thanks to Askar for his encouragement while developing prefix-free graphs.
    Thanks to Travis for introducing us to the advanced stringology.
    Thanks to Adrian for discussing some ideas from this paper with us.
\end{acknowledgments}

@@

\appendix

\section{Online Resources}

The sources for the prefix-free graphs are available via
\begin{itemize}
\item \href{https://github.com/andynet/pfg}{GitHub}.
\end{itemize}

>>>>

\section{Online Resources}
The sources for the prefix-free graphs are available via
\begin{itemize}
\item \href{https://github.com/andynet/pfg}{https://github.com/andynet/pfg}.
\end{itemize}

\begin{acknowledgments}
This research was funded by a grant from the European Union’s Horizon 2020 research and innovation programme under the Marie Skłodowska-Curie grant agreement No 956229 (ALPACA) and by a grant from Slovak Research Grant Agency VEGA 1/0538/22.
\end{acknowledgments}

@@

### 2023-06-26 manuscript/pfg.tex
finished

In prefix-free graphs, we partition each sequence in a given pangenome into \emph{segment}s.
These segments form nodes of the prefix-free graph, and their adjacencies in the original sequences constitute edges.
The sequences are represented as \emph{path}s in the graph.

@@

To create a prefix-free graph from a given pangenome $P$, we define a set of trigger words $T$, where each trigger is a string of length $k$.
For this set $T$, we build an Aho-Corasick automaton \cite{aho1975efficient}.
Then, for each sequence in pangenome $P$, we append $k$ sentinel characters and iterate over such modified sequence, searching for matches with set $T$ using the automaton.
Each time we encounter a trigger word, we recognize a new segment from the start of the previous trigger word to the end of the current trigger word.
If the segment's sequence was not yet observed during the scan, we add it to the set of segments and assign a unique ID.
Each time we append the corresponding ID to the path representing the original sequence.

>>>>

In prefix-free graphs, we partition each sequence of a given pangenome into \emph{segment}s.
These segments form nodes of the prefix-free graph, and their adjacencies in the original sequences constitute edges.
The sequences are represented as \emph{path}s in the graph.

@@

To create a prefix-free graph from a given pangenome $P$, we define a set of trigger words $T$, where each trigger word is a string of length $k$.
For this set $T$, we build an Aho-Corasick automaton \cite{aho1975efficient}.
Then, for each sequence in pangenome $P$, we append $k$ sentinel characters and iterate over such modified sequence, searching for matches with set $T$ using the automaton.
Each time we encounter a trigger word, we recognize a new segment from the start of the previous trigger word to the end of the current trigger word.
If the segment's sequence was not yet observed during the scan, we add it to the set of segments and assign a unique ID.
Each time we append the corresponding ID to the path representing the original sequence.

### 2023-06-26 manuscript/sa_construction.tex
finished

Here, we show another crucial advantage of prefix-free graphs.
Although they do not offer any improvement in theoretical guarantees in the worst case, in practice, they often represent the pangenome in a substantially smaller space and allow us to generate the suffix array values one by one, possibly using them directly in subsequential computation or storing them in compressed form.
This iteration can be done without ever expanding the pangenome to its textual representation in space proportional to the sum of segment lengths and the sum of path lengths.

\subsection{Iterator preparation}
To prepare the iterator of a suffix array of the pangenome from a prefix-free graph, we need to create several data structures.
First, we concatenate all the segments into a single string using a separator $\#$ and append a sentinel $\$$.
We will call this concatenation \emph{segment join}.
An example of a segment join is in Figure \ref{fig:ids_and_positions}.

@@

Next, for each suffix of the segment join, we need to calculate the corresponding segment ID and position values.
The value segment ID represents in what segment the current suffix starts, and the value segment position represents at what position in that particular segment the current suffix starts.
These arrays can be computed using an inverse permutation of a suffix array of a segment join in linear time.
To illustrate the procedure, consider the segment join of our running example from Figure \ref{fig:ids_and_positions}.
Each position of the join can be assigned a segment ID and a position in the current segment by linearly scanning the segment join and incrementing the ID and position accordingly.
Then, applying the inverse permutation of a suffix array to these arrays changes the order of computed values in correspondence to the sorted suffixes.
Table \ref{tab:suffix} shows the resulting suffix array, LCP array,
segment ID array and segment position array.

@@

In Table \ref{tab:suffix}, one row can represent multiple positions of the
pangenome.
To identify these positions, we store some additional information in Table \ref{tab:segment}.
For each segment, we store its length, starting positions in the pangenome and the rank of the right context of these positions.
To calculate the starting positions and the ranks of the right contexts, we use a \emph{path join}.
Similarly to segment join, a path join is a concatenation with delimiters $\#$ and sentinel $\$$, but now constructed by concatenating the paths.
An example of a path join for our running example is in Figure \ref{fig:path_join}.
Starting positions can be calculated by cumulatively summing the lengths
of the segments in path join and subtracting the overlaps.

@@

In the third case, the position is inside the trigger word or the sentinels appended during the graph construction.
The positions inside the trigger words are represented twice in the suffix table, once at the end of a segment and a second time at the beginning of the following segment in the pangenome.
These ending positions can violate the prefix-free property of the segment suffixes and, therefore, can be sorted incorrectly.
Skipping through these positions ensures the prefix-free property for the rest of the suffixes and also avoids double reporting.
This choice also plays nicely with the previous choice of appending $k$ sentinels during the construction of a prefix-free graph, as these positions will not get reported either.
In conclusion, if the length of a current segment suffix is smaller or equal to the size of the trigger words $k$, we skip the row as in the previous cases.

@@

As an example, consider the row $13$ in the Suffix Table (\ref{tab:suffix}), the first row yielding a SA value.
Its segment ID is $0$, and from the Segment Table (\ref{tab:segment}), we see only one occurrence of segment $0$ with starting position $9$ in the pangenome.
The offset from the start of a segment $\texttt{pos}[13]$ is $0$.
Summing these two values, we get the first value of a suffix array $9 + 0 = 9$ corresponding to the lexicographically smallest suffix $P[9..] = \texttt{ACACT}$.

The second, slightly more complex case is a singleton block representing a segment suffix with several occurrences in the pangenome.
In this case, we must report as many suffix array values as the number of occurrences.
Because the starting segment positions in the Segment Table (\ref{tab:segment}) are sorted based on their right context rank, we can iterate through these starting positions and apply Equation \ref{eq:sa_value} to each of them.

As an example, consider the row $14$ in the Suffix Table (\ref{tab:suffix}).
This suffix occurs twice in the pangenome in segments starting at positions $15$ and $1$.
Since the offset from the start of a segment $\texttt{pos}[14]$ is $0$, we report a suffix array values $15 + 0 = 15$ and $1 + 0 = 1$, corresponding to the suffixes $P[15..] = \texttt{ACGACT}$ and $P[1..] = \texttt{ACGTACT}$.

@@

As an example, consider the block of rows $[20..21]$ in the Suffix Table (\ref{tab:suffix}).
The relevant segment IDs are $0$ and $3$, with segment starts at positions $9$, $8$, $14$ and $0$.
The right context ranks from smallest to highest are $4, 5, 6, 9$ with the corresponding segment starts $8, 14, 0, 9$.
Applying Equation \ref{eq:sa_value} to these segment starts yields a suffix array values $8 + 0 = 8$, $14 + 0 = 14$, $0 + 0 = 0$ and $9 + 1 = 10$, representing pangenome suffixes $P[8..] = \texttt{CACACT}$, $P[14..] = \texttt{CACGACT}$, $P[0..] = \texttt{CACGTACT}$, and $P[10..] = \texttt{CACT}$.

>>>>

Here, we show another crucial advantage of prefix-free graphs.
Although they do not offer any improvement of theoretical guarantees in the worst case, in practice, they often represent the pangenome in a substantially smaller space and allow us to generate the suffix array values one by one, possibly using the values directly in subsequential computation or storing them in compressed form.
This iteration can be done without ever expanding the pangenome to its textual representation in space proportional to the sum of segment lengths and the sum of path lengths.

\subsection{Iterator preparation}
To prepare the iterator of a suffix array of the pangenome from a prefix-free graph, we need to create several data structures.
First, we concatenate all the segments into a single string using a separator $\texttt{\#}$ and append a sentinel $\texttt{\$}$.
We will call this concatenation \emph{segment join}.
An example of a segment join is in Figure \ref{fig:ids_and_positions}.

@@

Next, for each suffix of the segment join, we need to calculate the corresponding segment ID and position values.
The value segment ID represents in what segment the current suffix starts, and the value segment position represents at what position in that particular segment the current suffix starts.
These arrays can be computed using an inverse permutation of a suffix array $\texttt{ISA}$ (Equation \ref{eq:isa}) of a segment join in linear time.

\begin{equation}
    \label{eq:isa}
    \texttt{ISA}[\texttt{SA}[i]] = i
\end{equation}

To illustrate the procedure, consider the segment join of our running example from Figure \ref{fig:ids_and_positions}.
Each position of the join can be assigned a segment ID and a position in the current segment by linearly scanning the segment join and incrementing the ID and position accordingly.
Then, applying the inverse permutation of a suffix array to these arrays changes the order of computed values in correspondence to the sorted suffixes.
The resulting suffix array, LCP array, segment ID array and segment positions array are stored in a \emph{suffix table} as shown in Table \ref{tab:suffix}.

@@

In the suffix table, one row can represent multiple positions of the pangenome.
To identify these positions, we store some additional information in a \emph{segment table}.
For each segment, we store its length, starting positions in the pangenome and ranks of the right contexts of these positions.
To calculate the starting positions and the ranks of the right contexts, we use a \emph{path join}.
Similarly to segment join, a path join is a concatenation with delimiters $\texttt{\#}$ and sentinel $\texttt{\$}$, but now constructed by concatenating the paths.
An example of a path join for our running example is in Figure \ref{fig:path_join}.
Then, starting positions can be calculated by cumulatively summing the lengths
of the segments in path join and subtracting the overlaps.

@@

In the third case, the position is inside the trigger word or the sentinels appended during the graph construction.
The positions inside the trigger words are represented twice in the suffix table, once at the end of a segment and a second time at the beginning of the following segment in the pangenome.
These ending positions can violate the prefix-free property of the segment suffixes and, therefore, can be sorted incorrectly.
Skipping through these positions ensures the prefix-free property for the rest of the suffixes and also avoids double reporting.
Therefore, if the length of a current segment suffix is smaller or equal to the size of the trigger words $k$, we skip the row as in the previous cases.
This choice also plays nicely with the previous choice of appending $k$ sentinels during the construction of a prefix-free graph, as these positions will not get reported either.

@@

As an example, consider the row $13$ in Table \ref{tab:suffix}, the first row yielding a SA value.
Its segment ID is $0$, and from Table \ref{tab:segment}, we see only one occurrence of segment $0$ with starting position $9$ in the pangenome.
The offset from the start of a segment $\texttt{pos}[13]$ is $0$.
Summing these two values, we get the first value of a suffix array $9 + 0 = 9$ corresponding to the lexicographically smallest suffix $P[9..] = \texttt{ACACT}$.

The second, slightly more complex case is a singleton block representing a segment suffix with several occurrences in the pangenome.
In this case, we must report as many suffix array values as the number of occurrences.
Because the starting segment positions in the segment table are sorted based on their right context rank, we can iterate through these starting positions and apply Equation \ref{eq:sa_value} to each of them.

As an example, consider the row $14$ in Table \ref{tab:suffix}.
This suffix occurs twice in the pangenome in segments starting at positions $15$ and $1$.
Since the offset from the start of a segment $\texttt{pos}[14]$ is $0$, we report a suffix array values $15 + 0 = 15$ and $1 + 0 = 1$, corresponding to the suffixes $P[15..] = \texttt{ACGACT}$ and $P[1..] = \texttt{ACGTACT}$.

@@

As an example, consider the block of rows $[20..21]$ in Table \ref{tab:suffix}.
The relevant segment IDs are $0$ and $3$, with segment starts at positions $9$, $8$, $14$ and $0$.
The right context ranks from smallest to highest are $4, 5, 6, 9$ with the corresponding segment starts $8, 14, 0, 9$.
Applying Equation \ref{eq:sa_value} to these segment starts yields a suffix array values $8 + 0 = 8$, $14 + 0 = 14$, $0 + 0 = 0$ and $9 + 1 = 10$, representing pangenome suffixes $P[8..] = \texttt{CACACT}$, $P[14..] = \texttt{CACGACT}$, $P[0..] = \texttt{CACGTACT}$, and $P[10..] = \texttt{CACT}$.
